---
title: Matched tumor/normal pairs--a use case for the GenomicDataCommons Bioconductor package
author: Sean Davis
date: '2017-03-04'
categories:
  - Software
  - Tutorial
tags:
  - Bioconductor
  - GenomicDataCommons
  - Software
  - Data Science
  - API
  - R
---

The NCI Genomic Data Commons ([GDC](https://gdc.nci.nih.gov)) is a reboot of the approach that NCI uses to manage and expose genomic and associated clinical and experimental metadata. I have been working on a [Bioconductor](https://bioconductor.org) package that interfaces with the [GDC API](https://gdc-api.nci.nih.gov) to provide search and data retrieval from within R. 

In the first of what will likely be a set of use cases for the [GenomicDataCommons](https://github.com/Bioconductor/GenomicDataCommons), I am going to address a question that came up on twitter from [\@sleight82](https://twitter.com/sleight82)

> [Asking for a non-tweeter: can you find matched control samples?](https://twitter.com/sleight82/status/837898737540198400)

{{ tweet 837898737540198400 }}

The answer is, "Yes." I am going to assume that what the "non-tweeter" friend meant was that he or she was interested in finding matched tumor/normal data related to, for example, gene expression, and that the interest is in a specific disease category or TCGA. So, I am going to answer the more specific question:

> Can you find matched primary tumor/solid tissue normal samples and associated RNA-Seq gene expression files from TCGA breast cancer cases?

```{r}
library(GenomicDataCommons)
```

The GDC API puts some constraints on what can be done directly. But, since we are working in R, we have a toolbox that allows us to build a solution out of component parts. The strategy that I am going to employ is a three-step approach

1. Find RNA-Seq gene expression files derived from "Solid Tissue Normal"
2. Find RNA-Seq gene expression files derived from "Primary Tumor"
3. Limit cases from #1 and #2 that have gene expression results from both normal and tumor.

In this code block, find all "HTSeq - Counts" files that are derived from "Solid Tissue Normal" in the project "TCGA-BRCA". I used a combination of `files() %>% select(available_fields('files') %>% results()` to get an overview of the data available in the `files()` endpoint, followed by `grep_fields()` and `available_values()` to determine how to build filters. 

```{r}
nl_ge_files = files() %>%
    filter(~   cases.samples.sample_type=='Solid Tissue Normal' &
               cases.project.project_id == 'TCGA-BRCA' &
               analysis.workflow_type == "HTSeq - Counts") %>%
    GenomicDataCommons::select(c(default_fields('files'),
                                 'cases.case_id',
                                 'cases.samples.sample_type')) %>%
    results_all() %>%
    as.data.frame()
```

And do the same, but now looking for gene expression from tumors.

```{r}
tm_ge_files = files() %>%
    filter(~   cases.samples.sample_type=='Primary Tumor' &
               cases.project.project_id == 'TCGA-BRCA' &
               analysis.workflow_type == "HTSeq - Counts") %>%
    GenomicDataCommons::select(c(default_fields('files'),
                                 'cases.case_id',
                                 'cases.samples.sample_type')) %>%
    results_all() %>%
    as.data.frame()
```

Now, we have two data frames describing the normal- and tumor-derived TCGA-BRCA gene expression files. Note that I asked the GDC to provide `cases.case_id` as part of the record using `select()`. By looking for the intersection of cases between these two sets of files, we can find cases that contain files derived from both tumor and normal tissue.

```{r}
matchedcases = intersect(tm_ge_files$cases.case_id,
                         nl_ge_files$cases.case_id)
# how many matched cases?
length(matchedcases)
```

We can now create a `data.frame` that contains file information for only the matched samples:

```{r }
matched_tn_ge_file_info = rbind(subset(nl_ge_files,cases.case_id %in% matchedcases),
                                subset(tm_ge_files,cases.case_id %in% matchedcases))
head(matched_tn_ge_file_info)
```

Since the gene expression data are not that big, we can use the GDC API to download the data files directly. I use the BiocParallel package to facilitate parallel downloads.

```{r eval=FALSE}
library(BiocParallel)
register(MulticoreParam())
destdir = tempdir() # use a temp directory for example, change for yourself
fnames = bplapply(as.character(matched_tn_ge_file_info$file_id),
                  function(file_id) {
                      gdcdata(file_id, progress = FALSE,
                              destination_dir = destdir,
                              overwrite = TRUE)
                  })
```

Now, `fnames` should be a list of file names that can be read into and processed with R.

```{r sessionInfo}
sessionInfo()
```
